---
title: "Cancer data visualization and analysis"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: 
    fig_height: 4
  pdf_document:
    latex_engine: xelatex
    fig_height: 4
---

A two-hour workshop for participants in STEP-UP summer program on cancer prevention and control. There are 16 students, mostly upper-division undergraduates.

## Updates:

**Add tinyurl for data dictionary**

### General flow to lesson:

+ At the beginning, look at the data (everyone opens on their own machine), do the tidy data explanation.
+ Add information about a data dictionary here, since the column names aren’t great (need to go to the original source to see what units are for these variables). Give them time to look over the data dictionary
+ Present one hypothesis, maybe see how language isolation predicts lung cancer incidence.
+ How do we go about addressing this?
    + Data visualization (an exercise for them in breakout rooms)
        + Do this visualization in the code
    + Statistical analyses
        + Discuss but don’t do
+ Ask them to generate hypotheses, picking one of the demographic variables to predict incidence; include term “disparity” in discussions.
+ Do the stats on language isolation and lung cancer
+ Do _another_ regression example, using earlier code.
+ Assign groups to test another hypothesis in breakout rooms
+ Show ice cream & sharks graph

### Oddjobs

+ Create a data dictionary, maybe even another csv file that can be displayed on the screen

#### Learning objectives

1. Describe “tidy data” principles of one observation per row, one data type per column
2. Explain how to show relationships in data with data visualization
3. Develop hypotheses to explain quantitative data patterns
4. Write code to visualize data and test hypotheses
5. Explain the difference between correlation & causation

<!--
1. Examine data to ensure quality
2. Develop hypotheses to explain trends
3. Visualize comparisons between two groups
4. Visualize relationships between two variables
5. Test relationships between variables
6. Understand the difference between correlation and causation
-->

## [DESCRIPTION OR MOTIVATION; 2-4 sentences that would be used for an announcement]

***

## Getting started

+ Start RStudio
+ Create a new project via File > New Project... > New Directory > New Project
+ Make `data` and `output` directories in the project
```{r create-directories, eval = FALSE}
dir.create("data")
dir.create("output")
```
+ Download data (come from [https://statecancerprofiles.cancer.gov/](https://statecancerprofiles.cancer.gov/))
    + These data are lung cancer data incidence rates for each state, along with some demographic data

```{r data-retrieve, eval = FALSE}
download.file(url = "tinyurl.com/cancer-data-csv", destfile = "data/cancer-data.csv")
```

Open the file in a spreadsheet program like Microsoft Excel, LibreOffice Sheets, or Google Sheets.

```{r load-data, echo = FALSE}
cancer_data <- read.csv(file = "data/cancer-data.csv")
```

Notice that we have `r ncol(cancer_data)` columns of data. These are in what is called 'tidy' format. That is because each row has the data for a single state and each column has only one kind of data in it.

Some of these column names are easy to interpret, others are not as useful. The names are very short and it makes them easy to do analyses with them, but they are not necessarily human friendly. So we need to also download the data dictionary that defines what each of those columns means.

```{r data-dictionary-retrieve, eval = FALSE}
download.file(url = "tinyurl.com/data-dictionary-csv", destfile = "data/data-dictionary.csv")
```

Open this file in your spreadsheet program and see what the columns represent.

Now, back in R, we can start working with the data. We need to load the data into R so we can do our analyses.

```{r data-loading}
cancer_data <- read.csv(file = "data/cancer-data.csv")
```

We can also take a quick glance at some of these data in r with the `head` and `summary` commands.

```{r data-head}
head(cancer_data)
```

```{r data-summary}
summary(cancer_data)
```

***

## So what?

But what can we actually do with these data? Well, a lot, really. It really depends on the question you are interested in asking. So one thing we might be interested in is how language isolation affects the rates of lung cancer. That is, do states with a higher percentage of households that have difficulty with the English language have higher rates of lung cancer? This might happen because public health efforts may only be offered in English, and thus not reach all groups in need.



## Exercise
+ Get in groups; take five minutes to come up with a hypothesis you can test with these data

***

## Comparing groups
### New script (reproducibility)

```{r t-test-script}
# Compare cancer incidence between sexes
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2019-06-05
```

### Read in data & visualize male v. female incidence
```{r compare-male-female}
cancer_data <- read.csv(file = "data/cancer-data.csv")
boxplot(cancer_data$female.lung, cancer_data$male.lung)
```

But the x-axis doesn't tell us what "1" and "2" correspond to.

```{r boxplot-axis-names}
boxplot(list(female = cancer_data$female.lung, male = cancer_data$male.lung))
```

### Run t-test
```{r sexes-t-test}
t.test(x = cancer_data$female.lung, y = cancer_data$male.lung)
```

***

## Exercise
+ Get in groups; take five minutes to draw how you might show relationship between variables; ideally comes from one of the hypotheses students generated

***

## Testing relationships
### New script
```{r regression-script}
# Compare cancer incidence between sexes
# Jeff Oliver
# jcoliver@email.arizona.edu
# 2019-06-05
```

### Plot data & describe hypothesis
```{r plot-two-variables}
cancer_data <- read.csv(file = "data/cancer-data.csv")
plot(x = cancer_data$income, y = cancer_data$male.lung)
```

Clean up the axis labels
```{r clean-plot}
plot(x = cancer_data$income, 
     y = cancer_data$male.lung,
     xlab = "Income (US $)",
     ylab = "Lung cancer incidence (males)")
```

### Test relationship
```{r linear-regression}
income.lm <- lm(cancer_data$male.lung ~ cancer_data$income)
summary(income.lm)
```

### Update plot
```{r update-regression-plot}
plot(x = cancer_data$income, 
     y = cancer_data$male.lung,
     xlab = "Income (US $)",
     ylab = "Lung cancer incidence (males)")
abline(income.lm)
```

### Correlation vs. Causation
Discuss:

```{r sharks, echo = FALSE}
ice.cream <- 1:12 + rnorm(n = 12, mean = 2)
shark.attacks <- round(5 * (1:12 + rnorm(n = 12, mean = 2)), digits = 0)
plot(x = ice.cream,
     y = shark.attacks,
     xlab = "Ice Cream Sales ($1,000)",
     ylab = "Shark Attacks")
abline(lm(shark.attacks ~ ice.cream))
```

### Some additional plotting options
```{r plot-both-sexes}
plot(x = cancer_data$income, 
     y = cancer_data$male.lung,
     xlab = "Income (US $)",
     ylab = "Lung cancer incidence",
     pch = 19,
     col = "orange")
points(x = cancer_data$income,
       y = cancer_data$female.lung,
       pch = 19,
       col = "forestgreen")
abline(income.lm, col = "orange", lty = 2)
abline(lm(female.lung ~ income, data = cancer_data), col = "forestgreen", lty = 2)
legend("topright",
       legend = c("Females", "Males"),
       col = c("forestgreen", "orange"),
       pch = 19,
       lty = 2)
```

***

## Additional resources

+ [resource one](url-one)
+ [resource two](url-two)
+ A [PDF version](https://jcoliver.github.io/learn-r/step-up-lesson.pdf) of this lesson

***

Questions?  e-mail me at <a href="mailto:jcoliver@email.arizona.edu">jcoliver@email.arizona.edu</a>.